---
title: "Medical Appointment Analysis"
author: "Nirmal Sai Swaroop Janapaneedi"
date: "26/05/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction
We'll try to evaluate the accuracy of the **Medical Appointment No-Show** dataset which is available on <a href="https://www.kaggle.com/joniarroba/noshowappointments">Kaggle.com</a>.

As part of this exercise, we'll also review the complete dataset and try to identify some pattern based on the data available. 

**This Report includes:**

* Data Analysis and Visualization
* Split the data into Train(90%) and Test(10%) set
* Training Models
* Compare Accuracy with Test data set
* Combining of all train models results into Ensemble to predict the Optimal accuracy by decrease any variance, if any.

**Intial Setup:**

* Loading required Libraries
* Creating dataset from CSV file
* Clean-up data
* Defining functions to plot graph

```{r warning=FALSE, message=FALSE, echo=TRUE, cache=TRUE}
#load Libraries if not available
if(!require(tidyverse)) 
  install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) 
  install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(summarytools)) 
  install.packages("summarytools", repos = "http://cran.us.r-project.org")
if(!require(gbm)) 
  install.packages("gbm", repos = "http://cran.us.r-project.org")
if(!require(devtools)) 
  install.packages("devtools", repos = "http://cran.us.r-project.org")

# Read the dataset from csv file using the read_csv function
ds_apptnoshow <- read_csv("https://raw.githubusercontent.com/varora9/EDX_Project/master/KaggleV2-May-2016.csv", 
                col_names = c("PatientId","AppointmentID","Gender","ScheduledDay","AppointmentDay","Age",
                "Neighbourhood","Scholarship","Hipertension","Diabetes","Alcoholism",
                "Handcap","SMS_received","No-show"))
# remove the duplicate header row
ds_apptnoshow <- ds_apptnoshow[-1,]

# Covert columns which has numeric data from chr into double. This is required for training the model.
ds_apptnoshow$Age = as.double(ds_apptnoshow$Age)
ds_apptnoshow$Scholarship = as.double(ds_apptnoshow$Scholarship)
ds_apptnoshow$Hipertension = as.double(ds_apptnoshow$Hipertension)
ds_apptnoshow$Diabetes = as.double(ds_apptnoshow$Diabetes)
ds_apptnoshow$Alcoholism = as.double(ds_apptnoshow$Alcoholism)
ds_apptnoshow$Handcap = as.double(ds_apptnoshow$Handcap)
ds_apptnoshow$SMS_received = as.double(ds_apptnoshow$SMS_received)

#Rename No-Show Column Name
names(ds_apptnoshow)[names(ds_apptnoshow) == 'No-show'] <- 'NoShow'

# Defining function to generate graph based on columnname
plotGraphbasedonColumnname <- function(title, colname){
  ds_apptnoshow %>% ggplot(aes(get(colname))) +
    geom_bar(stat='count',colour="darkgreen", fill="lightgreen") +
    geom_text(aes(label = scales::percent((..count..)/sum(..count..))), 
              stat = "count", vjust = -0.5, colour="black" ) +
    labs(title = title,
         x = colname, 
         y = "Count")
}


# Defining function to generate graph based on columnname in facet
plotGraphbasedonNoShow <- function(title, colname){
  cols<- c('lightgreen','darkred')
  ds_apptnoshow %>% ggplot(aes(NoShow, fill=NoShow)) +
    geom_bar(stat='count',colour="darkgreen") +
    facet_grid(~get(colname)) +
    geom_text(aes(label = 
    scales::percent((..count..)/sapply(PANEL, 
                              FUN=function(x) sum(count[PANEL == x])))),
            stat = "count", vjust = -0.5, hjust= 0.5, colour="black" ) +
    scale_fill_manual(name="Bars",values=cols) +
    labs(title = title,
         x = "No-Show", 
         y = "Count")
}
```

# 2. Data Analysis and Visualization

**Data Dictionary:**

* PatientId - Identification of a patient
* AppointmentID - Identification of each appointment
* Gender - Male or Female .
* ScheduledDay - The day of the actual appointment, when they have to visit the doctor.
* AppointmentDay - The day someone registered the appointment, this is before appointment of course.
* Age - Patient's Age. 
* Scholarship - True of False . Data is stored in binary format (1,0)
* Hipertension - True or False. Data is stored in binary format (1,0).
* Alcoholism - True or False. Data is stored in binary format (1,0).
* Handcap - Data is stored in the range of 1 to 4.
* SMS_received - 1 or more messages sent to the patient.
* No-show - Yes or No. Yes = Didn't show-up, No = Show-up.

**Observation** - Scholarship is a broad topic, consider reading this <a href = "https://en.wikipedia.org/wiki/Bolsa_Fam%C3%ADlia">article</a>.

**a. Quick glance on the Appointment No-Show Dataset**

```{r echo=TRUE, cache=TRUE}
#Print Top 6 rows of the dataset
head(ds_apptnoshow)
```

**b. Quick glance of the Datset's Summary**

```{r echo=TRUE, cache=TRUE}
#summary of Dataset
summary(ds_apptnoshow)
```

**c. Medical Appointment No-Show dataset contain `r dim(ds_apptnoshow)[1]` rows and `r dim(ds_apptnoshow)[2]` columns.**

**d. Total count of unique Patients - `r length(unique(ds_apptnoshow$PatientId))`**


**e. Data segregation based on Gender. Females patients are significantly higher than Male patients.** 

```{r echo=TRUE, cache=TRUE}
# Print Dataset no. based on Gender
ds_apptnoshow %>% group_by(Gender) %>% summarize(Gender_Count= n()) 

# plot graph to depict Gender %
ds_apptnoshow %>% ggplot(aes(Gender)) +
  geom_bar(stat='count', fill="lightgreen", color="darkgreen") +
  geom_text(aes(label = scales::percent((..count..)/sum(..count..))), 
            stat = "count", vjust = -0.45, hjust = 0.5, colour="black" ) +
  labs(title = "Gender Ratio",
       x = "Gender", 
       y = "Count")
```

**f. Multiple patients have taken appointment multiple times and here is quick glance of the top 10 appointments** Maximum appointments taken by a Patient is **88**.

```{r echo=TRUE, cache=TRUE}
# No. of appointment taken by same patients
AppointCnt_Per_Patient <- ds_apptnoshow %>% 
  group_by(PatientId) %>% 
  summarize(AppointCnt = n())

# Top 10 Rows based on highest appointment count
top_n(AppointCnt_Per_Patient %>% arrange(desc(AppointCnt)),10)
```

**g. 3.2% Age data in dataset is incorrect where Age is documented as less than or equal to 0.**
Here is the quick review of the data.

```{r echo=TRUE, cache=TRUE}
# Review Age Data and fount incorrect Age data based on Gender.
ds_apptnoshow %>% filter(Age <= 0) %>% 
  group_by(Gender) %>% summarize(AgeErrorCount = n() )
```  

**h. Only 32% Patients received SMS of their appointments and out of that 28% patients didn't Show-Up.**

**Recommendation:** Doctor's office must need to look into this % to improve **No-Show** appointment rate.

```{r echo=TRUE, cache=TRUE}
# SMS Received by Patients of the appointments in Percentage
# 0 = No, 1 = Yes
summarytools::freq(ds_apptnoshow$SMS_received, order = "freq")[1:2, 4]

#Plot a graph to depict the no. of patients and No-Show by SMS Received
plotGraphbasedonColumnname("No. of Patients received SMS","SMS_received")
plotGraphbasedonNoShow("Patients received SMS with Show/No-Show %", "SMS_received")
```

**i. ~20% of the Patients didn't Show-Up post taking the appointments.** Probably, reminders would have helped them.

```{r echo=TRUE, cache=TRUE}
# No Show Frequency in Percentage
# 0 = Show Up, 1 = Didn't Show up
summarytools::freq(ds_apptnoshow$NoShow, order = "freq")[1:2, 4]
```

**j. Maximum Appointments observed on Wednesdays, followed by Tuesdays and minimal on Saturdays**.

```{r echo=TRUE, cache=TRUE}
# Adding Appointment Day in the main Dataset
ds_apptnoshow <- ds_apptnoshow %>% 
  mutate(apptdayoftheweek = weekdays(as.Date(AppointmentDay)), 
         apptday = as.numeric(format(AppointmentDay, format = "%u"))) 

# Appointment Count on day wise
ds_apptnoshow %>% 
  group_by(apptdayoftheweek, apptday) %>% 
  summarize(daywiseAppts=n()) %>% arrange(apptday) %>% 
  select(apptdayoftheweek, daywiseAppts)

#Plot Appointment Graph Day wise with % scale
ds_apptnoshow %>%
  ggplot(aes(apptdayoftheweek)) +
  geom_bar(stat='count',fill="lightgreen", color="darkgreen") +
  geom_text(aes(label = scales::percent((..count..)/sum(..count..))),
            stat = "count", vjust = -0.25, colour="black" ) +
  labs(title = "Daywise Appointments in % scale",
       x = "Week Days", 
       y = "Appointment Count")
```  

**k. % of No-Shows observed same on Mondays and Fridays but total count of Patients on Mondays are higher than Fridays.** 

**l. Overall percentage of No-Shows throughout week days is 19% or above.**

**m. No-Shows % on Saturdays is higher than other days but the Patients count is significantly low.**

```{r echo=TRUE, cache=TRUE}
# Appointment on day wise with Show/No-Show
plotGraphbasedonNoShow("Appointment on daywise with Show/No-Show ", "apptdayoftheweek")
```

```{r echo=TRUE, cache=TRUE}
# Data Clean-up - Set Age to 0 where it is stored as <0
ds_apptnoshow[ds_apptnoshow$Age <0,]$Age <- 0

# Create the Age Buckets to depict Age Pattern
ds_apptnoshow$AgeBuckets <- cut(ds_apptnoshow$Age, c(seq(0, 120, 10)), 
                      include.lowest=TRUE, right = FALSE)
```

**n. Maximum no. of Patients observed in the Age Bucket of 0-10 followed by 50-60.** 

```{r echo=TRUE, cache=TRUE}
#Plot a graph to depict the no. of patients and No-Show by Age Buckets
ds_apptnoshow %>% 
        group_by(AgeBuckets) %>% summarise(Total=n()) %>%
        ggplot(aes(AgeBuckets, Total)) +
        geom_bar(stat='identity',colour="darkgreen", fill="lightgreen") + 
        geom_text(aes(label = Total),  vjust = -0.45) + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1))  +
        labs(title = "No. of Appointments as per Age Buckets",
             x = "AgeBuckets", 
             y = "Count")
```

**o. Maximum % of No-Show Patients observed in the Age Bucket of 110-120 but the no.of patients in that is significantly low hence we can ignore that No-Show ratio.**

**Patients in age bucket of 10-20 has maximum rate of No-Show followed by 20-30.**

```{r echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
    
    ds_apptnoshow  %>% 
    group_by(AgeBuckets) %>% summarise(percentNoShow=round(sum(NoShow=='Yes')/n(),3)) %>%
    ggplot(aes(AgeBuckets, percentNoShow)) +
    geom_bar(stat='identity',colour="darkgreen", fill="lightgreen") + 
    geom_text(aes(label = paste(percentNoShow*100, "%")),  vjust = -0.45) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))  +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "% No-Show by Age Bucket",
         x = "AgeBuckets", 
         y = "% No-Show") +
    geom_hline(yintercept = mean(ds_apptnoshow$NoShow=='Yes'), col='darkred', size=1)
```

**p. 10% of the total Patients count has taken social aid from government for their illness expense and out of the that 24% of the Patients didn't Show-Up as per their Scheduled Appointments.**

```{r echo=TRUE, cache=TRUE}
#Plot a graph to depict the no. of patients and No-Show by Scholarship
plotGraphbasedonColumnname("No. of Patients with Scholarship", "Scholarship")
plotGraphbasedonNoShow("Scholarship Patients with Show/No-Show %","Scholarship")
```

**q. 20% of the total Patients observed with Hipertension and out of the that 17% of the Patients didn't Show-Up as per their Scheduled Appointments.**

```{r echo=TRUE, cache=TRUE}
#Plot a graph to depict the no. of patients and No-Show by Hipertension
plotGraphbasedonColumnname("No. of Patients with Hipertension","Hipertension")
plotGraphbasedonNoShow("Hipertension Patients with Show/No-Show %","Hipertension")
```

**r. 7% of the total Patients observed with Diabetes and out of the that 18% of the Patients didn't Show-Up as per their Scheduled Appointments.**

```{r echo=TRUE, cache=TRUE}
#Plot a graph to depict the no. of patients and No-Show by Diabetes
plotGraphbasedonColumnname("No. of Patients with Diabetes","Diabetes")
plotGraphbasedonNoShow("Diabetes Patients with Show/No-Show %","Diabetes")
```

**s. 3% of the total Patients observed with Acholism and out of the that 20% of the Patients didn't Show-Up as per their Scheduled Appointments.**

```{r echo=TRUE, cache=TRUE}
#Plot a graph to depict the no. of patients and No-Show by Alcoholism
plotGraphbasedonColumnname("No. of Patients with Alcoholism","Alcoholism") 
plotGraphbasedonNoShow("Alcoholism Patients with Show/No-Show %","Alcoholism")
```

**t. 2% of the total Patients observed with Handicap and average 23% of the Patients didn't Show-Up as per their Scheduled Appointments.**

```{r echo=TRUE, cache=TRUE}
#Plot a graph to depict the no. of patients and No-Show by Handcap
plotGraphbasedonColumnname("No. of Patients with Handcap", "Handcap")
plotGraphbasedonNoShow("Handicap Patients with Show/No-Show %","Handcap")
```

```{r echo=TRUE, cache=TRUE}
#Data Cleanup
# Converting No-Show data in binary format in column NoShow; 1 = Yes (Didn't Show-Up), 
#0 = No (Show-Up)
ds_apptnoshow$NoShow <- ifelse(ds_apptnoshow$NoShow =='Yes', 1, 0)
ds_apptnoshow$NoShow <- as.factor(ds_apptnoshow$NoShow)

# Removing some columns (non relevant) from dataset to shorten the 
# overall execution time to train the data.
ds_apptnoshow <- ds_apptnoshow %>% select(-PatientId,-AppointmentID,
                          -ScheduledDay,-AppointmentDay, 
                          -Age, -Neighbourhood,  apptdayoftheweek, 
                          -AgeBuckets, -apptdayoftheweek, -apptday)
```

## Quick view of the Dataset columns post cleanup.

```{r echo=TRUE, cache=TRUE}
names(ds_apptnoshow)
```

**Top 6 rows of the Dataset post cleanup**

```{r echo=TRUE, cache=TRUE}
head(ds_apptnoshow)
```

# 3. Training Models
 
**Note: Execution of Models takes approx 15-20 mins.**

**Splitting Appointment NoShow Dataset into Train (90%) / Test (10%) dataset**

```{r echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
set.seed(1)
test_index <- createDataPartition(ds_apptnoshow$NoShow, times = 1, p = 0.1, list = FALSE)
train <- ds_apptnoshow[-test_index,]
test <- ds_apptnoshow[test_index,]
```

  * Train dataset contains **`r dim(train)[1]`** rows and **`r dim(train)[2]`** columns.

  * Test dataset contains **`r dim(test)[1]`** rows and **`r dim(test)[2]`** columns.

**Model 1: Latent Dirichlet Allocation (LDA) Model**

Linear Discriminant Analysis (LDA) is a dimensionality reduction technique which is commonly used for the supervised classification problems.

```{r echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
set.seed(1)
#set.seed(1, sample.kind = "Rounding")    # if using R 3.6 or later
#Linear Discriminant Analysis (LDA)
train_lda <- train(NoShow ~ ., data=train, method = "lda")
#to display Accuracy of the trained dataset
train_lda$results
#predict Accuracy by comparing it with test dataset
lda_preds <- predict(train_lda, test)
test_accuracy <- mean(lda_preds == test$NoShow)
#Add a result in Model Result table
model_results <- tibble(Models="LDA", 
                Trained_Set_Accuracy = max(train_lda$results$Accuracy), 
                Test_Set_Accuracy = test_accuracy)

```

**Model 2: Quadratic Discriminant Analysis (QDA) Model**

A quadratic classifier is used in machine learning and statistical classification to separate measurements of two or more classes of objects or events by a quadric surface. 

```{r echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
set.seed(1)
#set.seed(1, sample.kind = "Rounding")    # if using R 3.6 or later
#Quadratic Discriminant Analysis (QDA)
train_qda <- train(NoShow ~ ., data=train, method = "qda")
#to display Accuracy of the trained dataset
train_qda$results
#predict Accuracy by comparing it with test dataset
qda_preds <- predict(train_qda, test)
test_accuracy <-mean(qda_preds == test$NoShow)
#Add a result in Model Result table
model_results <- bind_rows(model_results,
                           tibble(Models="QDA", 
                          Trained_Set_Accuracy = max(train_qda$results$Accuracy), 
                          Test_Set_Accuracy = test_accuracy))
```

**Model 3: Linear Regression Model (GLM) Model**

Linear Regression is a machine learning algorithm based on supervised learning. It performs a regression task. Regression models a target prediction value based on independent variables. It is mostly used for finding out the relationship between variables and forecasting. Different regression models differ based on – the kind of relationship between dependent and independent variables, they are considering and the number of independent variables being used.

```{r echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
set.seed(1)
#set.seed(1, sample.kind = "Rounding")    # if using R 3.6 or later
#Linear Regression Model
train_glm <- train(NoShow ~ ., data=train, method = "glm")
#to display Accuracy of the trained dataset
train_glm$results
#predict Accuracy by comparing it with test dataset
glm_preds <- predict(train_glm, test)
test_accuracy <- mean(glm_preds == test$NoShow)
#Add a result in Model Result table
model_results <- bind_rows(model_results,
                           tibble(Models="GLM", 
                          Trained_Set_Accuracy = max(train_glm$results$Accuracy), 
                          Test_Set_Accuracy = test_accuracy))
```

**Model 4: Gradient Boosting Machine Model (GBM) Model**

Gradient boosting is a machine learning technique for regression and classification problems, which produces a prediction model in the form of an ensemble of weak prediction models, typically decision trees.

```{r echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
set.seed(1)
#set.seed(1, sample.kind = "Rounding")    # if using R 3.6 or later
# gradient boosting machine
train_gbm <- train(NoShow~., data=train, method="gbm", verbose=FALSE)
#to display Accuracy of the trained dataset
train_gbm$results

#predict Accuracy by comparing it with test dataset
gbm_preds <- predict(train_gbm, test)
test_accuracy <- mean(gbm_preds == test$NoShow)
#Add a result in Model Result table
model_results <- bind_rows(model_results,
                           tibble(Models="GBM", 
                          Trained_Set_Accuracy = max(train_gbm$results$Accuracy), 
                          Test_Set_Accuracy = test_accuracy))
```

**Model 5: TREE (RPART) Model**

Decision tree learning is one of the predictive modeling approaches used in statistics, data mining and machine learning. It uses a decision tree (as a predictive model) to go from observations about an item  to conclusions about the item's target value.

```{r echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
# Convert all objects of train dataset into factor to train the data with tree model
train$Gender <- as.factor(train$Gender)
train$Scholarship <- as.factor(train$Scholarship)
train$Hipertension <- as.factor(train$Hipertension)
train$Diabetes <- as.factor(train$Diabetes)
train$Alcoholism <- as.factor(train$Alcoholism)
train$Handcap <- as.factor(train$Handcap)
train$SMS_received <- as.factor(train$SMS_received)
train$NoShow <- as.factor(train$NoShow)

# Convert all objects of test dataset into factor for predicting a tree value
test$Gender <- as.factor(test$Gender)
test$Scholarship <- as.factor(test$Scholarship)
test$Hipertension <- as.factor(test$Hipertension)
test$Diabetes <- as.factor(test$Diabetes)
test$Alcoholism <- as.factor(test$Alcoholism)
test$Handcap <- as.factor(test$Handcap)
test$SMS_received <- as.factor(test$SMS_received)
test$NoShow <- as.factor(test$NoShow)

set.seed(1)
#set.seed(1, sample.kind = "Rounding")    # if using R 3.6 or later
# Tree Based Model
train_rpart <- train(NoShow ~ ., data=train, method = "rpart")
#to display Accuracy of the trained dataset
train_rpart$results


#predict Accuracy by comparing it with test dataset
rpart_preds <- predict(train_rpart, test)
test_accuracy <- mean(rpart_preds == test$NoShow)
#Add a result in Model Result table
model_results <- bind_rows(model_results,
                           tibble(Models="RPART", 
                          Trained_Set_Accuracy = max(train_rpart$results$Accuracy), 
                          Test_Set_Accuracy = test_accuracy))
```

# 4. Result

**Summary of all above models with Accuracy on the Trained and Test dataset.** 

```{r echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
#to display consolidated accuracy of trained and test data set
model_results %>% knitr::kable()
```

**Ensemble method**

In statistics and machine learning, ensemble methods use multiple learning algorithms to obtain better predictive performance than could be obtained from any of the constituent learning algorithms alone.

```{r echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
#Using Ensemble method to combine the predicted values to decrease variance
ensemble <- cbind(glm = glm_preds == "0", lda = lda_preds == "0", 
                  qda = qda_preds == "0", gbm = gbm_preds == "0", 
                  rpart = rpart_preds =="0")

#predict Accuracy by comparing it with test dataset
ensemble_preds <- ifelse(rowMeans(ensemble) > 0.5, "0", "1")
mean(ensemble_preds == test$NoShow)
```

# 5. Conclusion

Average accuracy acheived is ~80% from all models but is not an excellent result. 

Doctors' office can decrease the No-Show rate by sending the SMS reminders to all of their patient and it can help in an improving the predictability of the patients.

Data seems to be limited and doesn't have enough information to improve the results.

Data can be improved by adding more weights/factors that may be able to help in improving the overall accuracy for No-Show.
